Extended Data

### 2021-06-06 doc/manuscript/Dow_manuscript_Nature_cd.Rmd
Create Dow_manuscript_Nature_cd.Rmd
Creating my own temp version of this to edit because i'm more comfortable having two versions to compare.

Grant Harley^5^

Justin Maxwell^6^

Ian McGregor^1,7^

@@

Alan Tepley^1,9^

@@

5. Department of Geography & Geological Sciences, University of Idaho, ID, USA

6. Department of Geography, Indiana University, Bloomington, Indiana, USA

@@

schematic_summary_caption <- "**Figure 1 | Summary of tree growth changes under warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing how each responds to warmer maximum temperatures during a 'critical temperature window', defined as that with the strongest control over $DOY_{25}$; (b) Variable definitions and summary responses to warmer spring temperatures at two temperate forests and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate increases and decreases, respectively, 'n.s.' indicates no signficant correlation, and 'mixed' indicates a mix of signficant and non-signficant correlations, often in different directions."

@@

DOY_Tsensitivity_caption <- "**Figure 3 | Response of the timing of stem growth phenology to spring temperatures.** The days of year on which 25%, 50% and 75% of annual growth were acheived ($DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively) declinde significantly with mean $T_{max}$ during the critical temperature window (CTW). For each CTW $T_{max}$, the posterior mean of the fitted day of year is represented by the solid line and 95% credible intervals are represented by bands. Relationsips for additional growth parameters (Fig. 1a) aare given in Extended Data Figure 4. (**DOYrmax will be dropped.** [GitHub Issue 23](https://github.com/EcoClimLab/growth_phenology/issues/23))" #needs attention!!!

@@

As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and annual CO~2~ sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014; @crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019].
However, less is known about how spring temperatures affect tree stem growth, which consumes only a modest portion of the carbon fixed annually[@anderson-teixeira_carbon_2021] but sequesters that carbon in wood that has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014] or as wood products[@profft_forest_2009].
Here, we show that in two temperate deciduous forests, warmer maximum spring temperatures consistently shifted woody growth milestones earlier but had mixed, weak effects on peak growing season length, maximum daily growth rates, and annual growth.
The latter finding is confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, which show that annual growth is by far more sensitive to temperatures during the peak growing season than in the spring.
These findings align with findings that annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length[@fu_maximum_2019], with observations that warm springs result in earlier leaf senescence[@zohner_ongoing_2019; @zani_increased_2020], and with evidence that forest CO~2~ sequestration is sink-limited[@zani_increased_2020].
They contrast, however, with the expectation *in most models[@zani_increased_2020]* that an earlier start to growing season would significantly increase the rate of biomass accumulation and thereby enhance the CO~2~ sink of temperate deciduous forests.
Rather, our findings imply that *the extra CO~2~ sequestered in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to the long-lived woody biomass C pool and is therefore unlikely to have a lasting impact on the forest C balance.*
Thus, warming spring temperatures are unlikely to increase the woody productivity or strengthen the carbon sink of temperate deciduous forests.

@@

In recent decades, warming temperatures have resulted in widespread phenological changes indicative of earlier arrival of spring[@parmesan_globally_2003, @ibanez_forecasting_2010,@jeong_phenology_2011,@friedl_tale_2014] and sometimes later autumn senesence[@ibanez_forecasting_2010,@jeong_phenology_2011; @keenan_net_2014].
In temperate deciduous forests, spring leaf-out has been occurring earlier[@ibanez_forecasting_2010, @jeong_phenology_2011], as has the beginning of the annual period of net CO~2~ sequestration[@keenan_net_2014].
Warmer spring temperatures and earlier leaf-out are also linked to earlier leaf senescence[@zohner_ongoing_2019;@keenan_timing_2015], apparently because of sink-limitation to CO~2~ sequestration[@zani_increased_2020].
However, warmer summer and autumn temperatures can delay leaf senescence[@zohner_ongoing_2019; @crabbe_extreme_2016; @dragoni_evidence_2011; @jeong_phenology_2011; @ibanez_forecasting_2010; @vitasse_responses_2009], and as a result warming may advance or delay autumn senescence [REFS].
The net result is that spring warming generally results in an increased growing season length, measured as the period over which trees have leaves[@jeong_phenology_2011;@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and the ecosystem is a net CO~2~ sink[REFS].
The expectation, is that as growing season length increases, net annual carbon sequestered will increase as well (*i,e*, net ecosystem exchange, $NEE$)[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016; @fu_maximum_2019], and [this expectation is represented in models](https://github.com/EcoClimLab/growth_phenology/issues/56) [@zani_increased_2020].
However, recent evidence that leaf senescence is induced when the seasonal carbon-sink capacities of trees are reached suggests that warmer spring temperatures and earlier leaf-out could advance autumn senescence and *will not necessarily lead to drastically increased C sequestration*[@zani_increased_2020].

We know very little about how the longest-lived component of fixed carbon, the woody growth of trees, is responding to warming spring temperatures. 
In fact, we know little about the stem-growth phenology of temperate deciduous species, with multi-year records of growth phenology having been published for only one site [@doranngeville_radial_inreview]. *(We need to make sure we're not missing anything.)* 
Tree-ring records, which can be used to examine correlations of annual growth to monthly temperatures but not to understand growth phenology, reveal that growth of temperate deciduous trees tends to be most sensitive to summer temperatures (REFS), with some hints that warmer springs may have a modest positive effect on growth (e.g., Refs.[@helcoski_growing_2019] [MORE]).
Thus, tree-ring records do not necessarily corroborate findings that warming spring temperatures lengthen the growing season and increase annual CO~2~ sequestration. 
Characterizing phenological responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.

@@

First, using dendrometer band measurements taken throughout the growing season at the Smithsonian Conservation Biology Institute (SCBI; Virginia; n = **146** trees over 10 years) and Harvard Forest (Massachusetts; n = **755** trees over six years), we fit a growth model[@mcmahon_general_2015] to determine the days of year where 25, 50, and 75% annual growth was achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$), and total annual growth ($\Delta DBH$; Fig. `r schematic_summary`). 
This analysis was performed separately for ring- and diffuse-porous species, which differ in growth phenology[@doranngeville_radial_inreview], and set in the context of leaf phenology at these sites.

@@

To understand how annual growth increments have responded to spring temperatures at the centennial scale, we analyzed tree-ring chronologies from our focal study sites  (`r EDT_species`) and across an additional `r n_cores_sites-2` (Fig. `r tree_rings`, `r EDF_map`).
For `r n_chronologies_SCBI` species at SCBI and `r n_chronologies_HF` species at Harvard Forest, ring width indices (RWI) rarely exhibited significant correlations to either $T_{max}$ or $T_{min}$ in the months corresponding to the CTWs identified here (Fig. `r tree_rings`, `r EDF_tree_rings_TMIN`).
Although correlations were rarely significant, there was a consistent pattern in responses to $T_{max}$ across all species at both sites:
responses to $T_{max}$ were negative or neutral in all months from February to August, with the exception of one spring month at each site (April at SCBI, May at Harvard Forest) where growth displayed a positive or neutral responses to $T_{max}$ (Fig. `r tree_rings`).
In these months, positive correlations to $T_{max}$ were statistically significant (at p=0.05) for one of nine ring-porous species at SCBI and one of two diffuse-porous species at each site.
Starting in May at SCBI and June at Harvard Forest, negative correlations of RWI to $T_{max}$ were predominant, being statistically significant in at least one growing season month for all `r n_chronologies_SCBI` species at SCBI and one of `r n_chronologies_HF` species at Harvard Forest. 
Patterns in $T_{min}$ were fairly similar to those of $T_{max}$, but weaker and more stochastic, with more (non-significant) correlations going in opposite directions (`r EDF_tree_rings_TMIN`).

@@

The second study site was Harvard Forest (42.5388° N, 72.1755° W, 340-368 m.a.s.l.), located near the central Massachusetts town of Petersham.
Harvard forest is a secondary forest, having re-established following agricultural use *(time frame)* and signficant hurricane damage in *YEAR*. 
Dominate species within the 35 ha ForestGEO plot are hemlock (*Tsuga spp.*), oak (*Quercus spp.*) and maple (*Acer spp.*). 
The climate is temperate continental with a mean annual precipitation of 1120 mm and temperatures  averaging -12° C in January and 19° C in July (D'Orangeville et al, ).

@@

Metal dendrometer bands installed on a total of `r n_dendro_trees` trees within the SCBI and Harvard Forest ForestGEO plots (`r  EDT_species`). 
Bands were placed on dominant species, including two diffuse- and two ring-porous species at SCBI and eight diffuse- and three ring-porous species at Harvard Forest.
Bands were measured with a digital caliper approximately every 1-2 weeks within the growing season from 2011-2020 at SCBI and 1998-2003 at Harvard Forest, respectively. 
The number of bands measured at each site fluctuated somewhat as trees were added or dropped (e.g., because of tree mortality) from the census. 
Across years, the number of bands sampled averaged *~135* (range: 99-145) at SCBI and  *~717* (range: *700-755*) at Harvard Forest. 
In total, our analysis included `r n_tree_years` tree-years (`r  EDT_species`).

@@

### Tree rings {-#Tree} 
Dominant tree species were cored at both sites following sampling designs that covered a broad range of $DBH$.
We analyzed records for the dominant ring- and diffuse- porous species at each site (`r  EDT_species`), excluding species with other xylem architectures (*Juglans nigra* L. at SCBI, *Tsuga Canadensis* (L.) Carrière at Harvard Forest)
At SCBI, cores were collected as described in @helcoski_growing_2019. 
Briefly, cores were taken from living or recently dead trees $\ge$ 10 cm $DBH$, including a randomly sampled subset of all dominant species within the ForestGEO plot [live trees in 2010-2011; @bourg_initial_2013] and trees found newly dead in annual tree mortality in 2016 and 2017 [@helcoski_growing_2019].
At Harvard Forest, ...(*Neil can write this easily, or KAT probably has the info*)
In total, we analyzed `r n_cores_focal_sites` cores from eleven species at SCBI and three species at Harvard Forest (`r  EDT_species`).

>>>>

Grant L. Harley^5^

Justin T. Maxwell^6^

Ian R. McGregor^1,7^

@@

Alan J. Tepley^1,9^

@@

5. Department of Geography & Geological Sciences, University of Idaho, ID 83844, USA

6. Department of Geography, Indiana University, Bloomington, IN 47405, USA

@@

schematic_summary_caption <- "**Figure 1 | Summary of tree growth changes under warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing how each responds to warmer maximum temperatures during a 'critical temperature window (CTW)', defined as that with the strongest control over $DOY_{25}$; (b) Variable definitions and summary responses to warmer spring temperatures at two temperate forests and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate increases and decreases, respectively, 'n.s.' indicates no signficant correlation, and 'mixed' indicates a mix of signficant and non-signficant correlations, often in different directions."

@@

DOY_Tsensitivity_caption <- "**Figure 3 | Response of the timing of stem growth phenology to spring temperatures.** The days of year on which 25%, 50% and 75% of annual growth were acheived ($DOY_{25}$, $DOY_{50}$, and $DOY_{75}$, respectively) declined significantly with mean $T_{max}$ during their respective critical temperature window (CTW). For each CTW $T_{max}$, the posterior mean of the fitted day of year is represented by the solid line and 95% credible intervals are represented by bands. Relationships for additional growth parameters (Fig. 1a) are given in Extended Data Figure 4. (**DOYrmax will be dropped.** [GitHub Issue 23](https://github.com/EcoClimLab/growth_phenology/issues/23))" #needs attention!!!

@@

As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and annual CO~2~ sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014; @crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019].
However, less is known about how spring temperatures affect tree stem growth, which consumes only a modest portion of the carbon fixed annually[@anderson-teixeira_carbon_2021] but sequesters that carbon in wood that has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014] or as wood products[@profft_forest_2009].
Here, we show that in two temperate deciduous forests, warmer maximum spring temperatures consistently shifted woody growth milestones earlier but had mixed, weak effects on peak growing season length, maximum daily growth rates, and annual growth.

*The latter finding is confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` deciduous forests across eastern North America, which show that annual growth is by far more sensitive and generally negatively impacted by temperatures during the peak growing season than in the spring.*
*These findings align with previous studies that suggest annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length[@fu_maximum_2019] and with observations that warm springs result in earlier leaf senescence because CO~2~ sequestration is sink limited [@zohner_ongoing_2019; @zani_increased_2020].* (David Orwig asked if this is true in both evergreen and deciduous forest. Combining the last two points into one sentence may better represent what Zani found. Remove starting from "The Latter" to here?)

They contrast, however, with the expectation *in most models[@zani_increased_2020]* that an earlier start to growing season would significantly increase the rate of biomass accumulation and thereby enhance the CO~2~ sink of temperate deciduous forests.
Rather, our findings imply that *the extra CO~2~ sequestered in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to the long-lived woody biomass C pool and is therefore unlikely to have a lasting impact on the forest C balance.* (Add a line suggeesting this may be because of water stress?)
Thus, warming spring temperatures are unlikely to increase the woody productivity or strengthen the carbon sink of temperate deciduous forests.

@@

Warming temperatures in recent decades have resulted in widespread phenological changes indicative of earlier arrival of spring[@parmesan_globally_2003, @ibanez_forecasting_2010,@jeong_phenology_2011,@friedl_tale_2014] and sometimes later autumn senesence[@ibanez_forecasting_2010,@jeong_phenology_2011; @keenan_net_2014].
In temperate deciduous forests, spring leaf-out has been occurring earlier[@ibanez_forecasting_2010, @jeong_phenology_2011], as has the beginning of the annual period of net CO~2~ sequestration[@keenan_net_2014].
Warmer spring temperatures and earlier leaf-out are also linked to earlier leaf senescence[@zohner_ongoing_2019;@keenan_timing_2015], apparently because of sink-limitation to CO~2~ sequestration[@zani_increased_2020].
However, warmer summer and autumn temperatures can delay leaf senescence[@zohner_ongoing_2019; @crabbe_extreme_2016; @dragoni_evidence_2011; @jeong_phenology_2011; @ibanez_forecasting_2010; @vitasse_responses_2009], and as a result warming may advance or delay autumn senescence [REFS].
The net result is that spring warming generally results in an increased growing season length, measured as the period over which trees have leaves[@jeong_phenology_2011;@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and the ecosystem is a net CO~2~ sink[REFS].
The expectation, is that as growing season length increases, net annual carbon sequestered will increase as well (*i,e*, net ecosystem exchange, $NEE$)[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016; @fu_maximum_2019], and [this expectation is represented in models](https://github.com/EcoClimLab/growth_phenology/issues/56) [@zani_increased_2020].
However, recent evidence that leaf senescence is induced when the seasonal carbon-sink capacities of trees are reached suggests that warmer spring temperatures and earlier leaf-out could advance autumn senescence and *will not necessarily lead to drastically increased C sequestration*[@zani_increased_2020].

We know very little about how the longest-lived component of fixed carbon, the woody growth of trees, is responding to warming spring temperatures. 
In fact, we know little about the stem-growth phenology of temperate deciduous species, with multi-year records of growth phenology having been published for only one site [@doranngeville_radial_inreview]. *(We need to make sure we're not missing anything.)* 
Tree-ring records, which can be used to examine correlations of annual growth to monthly temperatures but not to understand growth phenology, reveal that growth of temperate deciduous trees tends to be most sensitive to summer temperatures (REFS), with some hints that warmer springs may have a modest positive effect on growth (e.g., Refs.[@helcoski_growing_2019] [MORE]).
Thus, tree-ring records do not necessarily support the suggestion that warming spring temperatures would increase annual CO~2~ sequestration. 
Characterizing phenological responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.(Loic suggests adding numbers to illustrate the importatnce of stem growth as a carbon sink.)

@@

First, using dendrometer band measurements taken throughout the growing season at the Smithsonian Conservation Biology Institute (SCBI; Virginia; n = **146** trees from 2011-2020) and Harvard Forest (Massachusetts; n = **755** trees from 1998-2003), we fit a growth model[@mcmahon_general_2015] to determine the days of year where 25, 50, and 75% annual growth was achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$), and total annual growth ($\Delta DBH$; Fig. `r schematic_summary`). 
This analysis was performed separately for ring- and diffuse-porous species, which differ in growth phenology[@doranngeville_radial_inreview], and set in the context of leaf phenology at these sites.

@@

To understand how annual growth increments have responded to spring temperatures at the centennial scale, we analyzed tree-ring chronologies from our focal study sites  (`r EDT_species`) and across an additional `r n_cores_sites-2` locations (Fig. `r tree_rings`, `r EDF_map`).
For `r n_chronologies_SCBI` species at SCBI and `r n_chronologies_HF` species at Harvard Forest, ring width indices (RWI) rarely exhibited significant correlations to either $T_{max}$ or $T_{min}$ in the months corresponding to the CTWs identified here (Fig. `r tree_rings`, `r EDF_tree_rings_TMIN`).
Although correlations were rarely significant, there was a consistent pattern in responses to $T_{max}$ across all species at both sites:
responses to $T_{max}$ were negative or neutral in all months from February to August, with the exception of one spring month at each site (April at SCBI, May at Harvard Forest) where growth displayed a positive or neutral responses to $T_{max}$ (Fig. `r tree_rings`).
In these months, positive correlations to $T_{max}$ were statistically significant (at p=0.05) for one of nine ring-porous species at SCBI and one of two diffuse-porous species at each site.
Starting in May at SCBI and June at Harvard Forest, negative correlations of RWI to $T_{max}$ were predominant, being statistically significant in at least one growing season month for all `r n_chronologies_SCBI` species at SCBI and one of `r n_chronologies_HF` species at Harvard Forest. 
Patterns in $T_{min}$ were fairly similar to those of $T_{max}$, but weaker and more stochastic, with more (non-significant) correlations going in opposite directions (`r EDF_tree_rings_TMIN`).

@@

The second study site was Harvard Forest (42.5388° N, 72.1755° W, 340-368 m.a.s.l.), located near the central Massachusetts town of Petersham.
Harvard forest is a secondary forest, having re-established following agricultural use *(time frame)* and significant hurricane damage in 1938. 
Dominate species within the 35 ha ForestGEO plot are hemlock (*Tsuga canadensis*), oak (*Quercus spp.*) and red maple (*Acer rubrum*). 
The climate is temperate continental with a mean annual precipitation of 1120 mm and temperatures  averaging -12° C in January and 19° C in July (D'Orangeville et al, ).

@@

Metal dendrometer bands were installed on a total of `r n_dendro_trees` trees within the SCBI and Harvard Forest ForestGEO plots (`r  EDT_species`). 
Bands were placed on dominant species, including two diffuse- and two ring-porous species at SCBI and eight diffuse- and three ring-porous species at Harvard Forest.
Bands were measured with a digital caliper approximately every 1-2 weeks within the growing season from 2011-2020 at SCBI and 1998-2003 at Harvard Forest, respectively. 
The number of bands measured at each site fluctuated somewhat as trees were added or dropped (e.g., because of tree mortality) from the census. 
Across years, the number of bands sampled averaged *~135* (range: 99-145) at SCBI and  *~717* (range: *700-755*) at Harvard Forest. 
In total, our analysis included `r n_tree_years` tree-years (`r  EDT_species`).

@@

### Tree rings {-#Tree} 
Dominant tree species were cored at both sites following sampling designs that covered a broad range of $DBH$.
We analyzed records for the dominant ring- and diffuse- porous species at each site (`r  EDT_species`), excluding species with other xylem architectures (*Juglans nigra* L. at SCBI, *Tsuga canadensis* (L.) Carrière at Harvard Forest)
At SCBI, cores were collected as described in @helcoski_growing_2019. 
Briefly, cores were taken from living or recently dead trees $\ge$ 10 cm $DBH$, including a randomly sampled subset of all dominant species within the ForestGEO plot [live trees in 2010-2011; @bourg_initial_2013] and trees found newly dead in annual tree mortality in 2016 and 2017 [@helcoski_growing_2019].
At Harvard Forest, ...(*Neil can write this easily, or KAT probably has the info*)
In total, we analyzed `r n_cores_focal_sites` cores from eleven species at SCBI and three species at Harvard Forest (`r  EDT_species`).
